Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument

### 2014-10-18 Papers/InProgress/LogicAndReligionCongress/Abstract.tex
updates for paper

\usepackage{url}
\usepackage{commands}
%
%
% THEOREM Environments (Examples)-----------------------------------------
%
 \newtheorem{thm}{Theorem}[section]
 \newtheorem{cor}[thm]{Corollary}
 \newtheorem{lem}[thm]{Lemma}
 \newtheorem{prop}[thm]{Proposition}
 \theoremstyle{definition}
 \newtheorem{defn}[thm]{Definition}
 \theoremstyle{remark}
 \newtheorem{rem}[thm]{Remark}
 \newtheorem*{ex}{Example}
 \numberwithin{equation}{section}

@@

\title[Modal Collapse]
 {Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument}

@@

\address{%
Department of Mathematics and Computer Science\\
Arnimallee 7 \\
Room 115 \\
14195 Berlin \\
Germany
}

\email{c.benzmueller@gmail.com}

@@

%----------Author 2
\author[Weber]{L. Weber}
\address{ 
ToDo \\
ToDo \\
ToDo Berlin \\
Germany
}
\email{leon.weber@fu-berlin.de}

@@

%\vspace*{-5em}

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument
\cite{GoedelNotes,ScottNotes} imply what is called the modal collapse
\cite{Sobel1987,SobelBook2004} (formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. God(x)$ as intended). This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{J30,ECAI-2014}, has led to strong criticism of the argument, and
it has therefore stimulated attempts to remedy the problem. Some
authors, including Hajek
\cite{Hajek_der_Mathematiker,Hajek_Magari_and_others}, suggest the use
of cautious instead of full comprehension principles in order to
escape the modal collapse. Others, including Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordahl \cite{Bjordahl}, have proposed to adapt G\"odel's axioms (in
particular, controversial axiom A1) and definitions.

We have studied --- with computer-support --- the emendations proposed
by Anderson and Bjordahl.  More precisely, we have formalized their
emendations within the proof assistant Isabelle/HOL \cite{Isabelle},
and we have then automated (different variations of) their emendations
with the higher-order reasoners Leo-II \cite{C26}, Satallax
\cite{brown2012satallax}, Metis \cite{Hurd03first-orderproof} and
Nitpick \cite{BN10}; these systems are available in Isabelle/HOL.  Our
formalizations\footnote{The formalizations are available in the
  subdirectories \url{Anderson} and \url{Bjordahl} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced and employed in our previous
work \cite{J30,ECAI-2014}. In this approach full comprehension is
naturally ``built-in'' since the underlying HOL supports
$\lambda$-abstraction. The assumption of full comprehension, up to our
understanding, is
in-line with Anderson and Bjordahl (?).

We summarize the main results of our experiments, and we start with
Anderson's emendation (cf. Appendix \ref{apx:Anderson}), which we have
analysed for different domain conditions. These variations were
motivated by various comments on Anderson's work in the literature
(maybe we should be more explicit).
\begin{itemize}
\item Constant domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_constant_domain.thy}} (the individual
  domains remain the same in all worlds): G\"odel's theorems C1, and
  T3 can easily be automated (in logics K and KB, respectively) ,
  G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic K4B, the latter already in K); a trivial countermodel (two
  worlds, one individual; this is simpler than
          the countermodel with two individuals and two worlds
          presented by Anderson, 1990) to MC generated by Nitpick (for all mentioned
  logics); this model also proves consistency.
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_varying_domain.thy}} (the domains for
  indivduals may vary from world to world, all other domains remain
  constant; different authors have pointed out that varying domain
  quantifiers should be used in Anderson's work; here we consistently
  used varying domain for all individual quantifiers; the varying
  domain quantifiers have been modeled by using an explicit
  ``Existence'' relation between indiviudals and worlds): we obtain
  the same results as  above
\item Mixed variant\footnote{Cf. Isabelle/HOL source files
    \url{Anderson/Anderson_mixed_domain_1/2.thy}} (varying domain
  quantifiers are used exclusively in the definitions of essence and
  NE; see the comments by Fitting): G\"odel's axioms A4 and A5 remain
  redundant (the former one in
  logic K4B, the latter already in K). The situation changes when only
  NE is modeled with a varying domain quantifier. Now, Nitpick reports
  a countermodel. 
\end{itemize}

Need to do Bjordahl \ldots.

@@

\begin{appendix}

% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
% \begin{itemize}
% \item[MC] Everything that is the case is so necessarily:
%   $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
% \item[MC'] Everything that is possible is necessary:
%   $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
% \item[MC''] Modalities collapse:
%   $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
% \end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
% \end{figure}

@@

@@

\section{Bjordahl's Alternative} \label{apx:Bjordahl}

% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
\begin{itemize}
\item[A1] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \phi [P(\neg \phi) \biimp \neg P(\phi)]}$$ 
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \phi \allq \psi [(P(\phi) \wedge \nec \allq x [\phi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified: 
  $$\hol{\allq \phi [P(\phi) \imp \pos \exq x \phi(x)]}$$ 
\item[D1] A \emph{God-like} being possesses all positive properties: 
  $$\hol{G(x) \biimp \forall \phi [P(\phi) \imp \phi(x)]}$$ 
\item[A3]  The property of being God-like is positive: 
  $$\hol{P(G)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive: 
  $$\hol{\allq \phi [P(\phi) \imp \Box \; P(\phi)]}$$ 
\item[D2] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\phi}{x} \biimp \phi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\phi(y) \imp \psi(y)))}$$ 
\item[T2]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[D3] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE(x) \biimp \allq \phi [\ess{\phi}{x} \imp \nec
  \exq y \phi(y)]}$$
\item[A5] Necessary existence is a positive property: $$\hol{P(\NE)}$$ 
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Frode's Alternative \cite{ToDo:Frode}.\label{fig:frode}} 
% \end{figure}

>>>>

\usepackage{url}
\usepackage{commands}

\usepackage[utf8]{inputenc}

%
%
% THEOREM Environments (Examples)-----------------------------------------
%
 \newtheorem{thm}{Theorem}[section]
 \newtheorem{cor}[thm]{Corollary}
 \newtheorem{lem}[thm]{Lemma}
 \newtheorem{prop}[thm]{Proposition}
 \theoremstyle{definition}
 \newtheorem{defn}[thm]{Definition}
 \theoremstyle{remark}
 \newtheorem{rem}[thm]{Remark}
 \newtheorem*{ex}{Example}
 \numberwithin{equation}{section}

@@

\title[Analysis of Modern Variants of G\"odel's Ontological Argument]
 {Computer-supported Analysis of Modern Variants of G\"odel's Ontological Argument}

@@

\address{%
Dep.~of Mathematics and Computer Science, Freie Universit\"at Berlin, Germany
}
\email{c.benzmueller@fu-berlin.com}

@@

%----------Author 2
\author[Weber]{L. Weber}
\address{ 
Dep.~of Mathematics and Computer Science, Freie Universit\"at Berlin, Germany
}
\email{leon.weber@fu-berlin.de}

@@

\vspace*{-5em}

%\section{The Modal Collapse}
The axioms in G\"odel's ontological argument 
\cite{GoedelNotes,ScottNotes} (cf. Appendix \ref{apx:Goedel}) imply what is called the modal collapse
\cite{Sobel1987,SobelBook2004} (formula $\varphi \rightarrow \Box
\varphi$, abbreviated as MC, holds for any formula $\varphi$ and not
just for $\exists x. God(x)$ as intended). This fact, which has
recently been confirmed with higher-order automated theorem provers
\cite{C40,J30}, has led to strong criticism of the argument, and
it has therefore stimulated attempts to remedy the problem. Some
authors, including Hajek
\cite{Hajek_der_Mathematiker_2002,Hajek_Magari_and_others_1996}, suggest the use
of cautious instead of full comprehension principles in order to
escape the modal collapse. Others, including Anderson
\cite{anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings}
and Bjordal \cite{bjordal99}, propose to slightly modify
G\"odel's axioms and definitions.

We have studied --- with computer-support --- the emendations proposed
by Anderson and Bjordal (thereby we have retained full comprehension).
More precisely, we have formalized their emendations within the proof
assistant Isabelle/HOL \cite{Isabelle}, and we have then automated
(different variations of) their emendations with the higher-order
reasoners Leo-II \cite{C26}, Satallax \cite{brown2012satallax}, Metis
\cite{Hurd03first-orderproof} and Nitpick \cite{Nitpick}; these systems
are available in Isabelle/HOL.  Our formalizations\footnote{The
  formalizations are available in the subdirectories \url{Anderson}
  and \url{Bjordal} at
  \url{https://github.com/FormalTheology/GoedelGod/blob/master/Formalizations/Isabelle/}.}
employ the embedding of higher-order modal logic (HOML) in classical
higher-order logic (HOL) as introduced and employed in previous
work \cite{C40,J30,J23}. In this approach full comprehension is
naturally ``built-in'' since the underlying HOL supports
$\lambda$-abstraction.

We summarize the main results of our experiments, and we start with
Anderson's emendation (cf. Appendix \ref{apx:Anderson}; @Bruno: I do
not yet address T2', L1' and L2' below; we could eventually include those as well?),
which we have analysed for different domain conditions. These
variations were motivated by various comments on Anderson's work in
the literature (@Bruno: maybe we should be more explicit about these
comments).
\begin{itemize}
\item Constant domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_constant_domain.thy}} (the individual
  domains remain unchanged for all worlds): G\"odel's theorems T1, C
  and T3' can be quickly automated (in logics K and KB, respectively),
  G\"odel's axioms A4 and A5 are proven redundant (the former one in
  logic K4B and the latter one already in K). Moreover, a trivial
  countermodel (consisting of two worlds and two individuals) to MC
  generated by Nitpick (for all mentioned logics); this model also
  proves consistency. 
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source file
    \url{Anderson/Anderson_varying_domain.thy}} (the domains for
  indivduals may vary from world to world, all other domains remain
  constant; various authors have pointed out that varying domain
  quantifiers should be used in Anderson's work; here we consistently
  used varying domain for all individual quantifiers; the modeling of
  the varying domain quantifiers employs an explicit ``Existence''
  relation between indiviudals and worlds as guard): In this setting we obtain the
  same results as above.
\item Mixed variant\footnote{Cf. Isabelle/HOL source files
    \url{Anderson/Anderson_mixed_domain_1/2.thy}} (varying domain
  quantifiers are used only in the definitions of essence and NE;
  cf.~Fitting comments to Anderson in
  \cite{fitting02:_types_tableaus_god}): Also in this setting we
  obtain the same results as above. However, if a varying domain
  quantifier is used only in the definition of NE, then the
  situation slightly changes. Now axiom A5 is no longer provable and
  a countermodel is reported by Nitpick. The remaining results are as before.

%   G\"odel's axioms A4 and A5
%   remain redundant (the former one in logic K4B, the latter already in
%   K). The situation changes when only NE is modeled with a varying
%   domain quantifier. Now, Nitpick reports a countermodel.
\end{itemize}

Our analysis of Bjordal emendation of G\"odel's argument (cf. Appendix \ref{apx:Bjordal}) has produced the following results:
\begin{itemize}
\item Constant domain semantics\footnote{Cf. Isabelle/HOL source files
    \url{Bjordal/Bjordal_A/B/C_constant_domain.thy}}: G\"odel's axiom
  A2, A3 can be quickly automatically derived in logic K from
  Bjordal's definition B:D. A4 can be proved in logic T
  (reflexivity). Proving G\"odel's D1 from B:D is possible in logic
  K4. Vice versa, for the proof that B:D follows from D1, A2, A3 and
  A4 the povers succeed already in logic K. Hence, Bjordal's lemma
  B:L1 holds in logic S4. The provers also show that theorem T3
  follows from B:D, B:A1 and B:A2 already in logic KB. Modal collapse
  does not follow in Bjordal's setting as Nitpick demonstrates with a
  countermodel (consisting of two worlds and one individual). @Bruno: we could eventually also
  also say that for MC1 we need a bigger countermodel?
\item Varying domain semantics\footnote{Cf. Isabelle/HOL source files
    \url{Bjordal/Bjordal_A/B/C_varying_domain.thy}}: The results are
  the same as above. 
\item @Bruno: we could also check the mixed case her; but there doesn't seem to be strong reason for doing so.
\end{itemize}

Summary (what else can we say here, feel free to add): Using our approach, the formalization and (partly) automated
analysis of different variants of Anderson's and Bjordal's emendations
of G\"odel's ontological argument has been surprisingly
straightforward. The provers confirmed the claimed results and in a
few cases they have even contributed some novel insights. The
weakening of the comprehension principles would clearly constitute
another interesting parameter for further experiments. However, this
seems hard to achieve in our approach, since full comprehension is
naturally built-in.

@@

\newpage
\begin{appendix}
\noindent @Bruno: I think we need more compact presentation of the proofs.
% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
% \begin{itemize}
% \item[MC] Everything that is the case is so necessarily:
%   $$\hol{\allq \phi [\phi \imp \nec \phi]}$$ 
% \item[MC'] Everything that is possible is necessary:
%   $$\hol{\allq \phi [\pos \phi \imp \nec \phi]}$$ 
% \item[MC''] Modalities collapse:
%   $$\hol{\allq \phi [\phi \biimp \pos \phi \biimp \nec \phi]}$$
% \end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Modal Collapse \cite{todo:Sobel1987,SobelBook}.\label{fig:collapse}} 
% \end{figure}

@@

\small

\section{Scott's version of G\"odel's ontological argment} \label{apx:Goedel} 
\begin{itemize}
\item[A1] Either a property or its negation is positive, but not
  both:
  $$\hol{\allq \varphi [P(\neg \varphi) \biimp \neg P(\varphi)]}$$ 
\item[A2] A property necessarily implied by a
  positive property is positive:
  $$\hol{\allq \varphi \allq \psi [(P(\varphi) \wedge \nec \allq x [\varphi(x)
  \imp \psi(x)]) \imp P(\psi)]}$$
\item[T1] Positive properties are possibly exemplified: 
  $$\hol{\allq \varphi [P(\varphi) \imp \pos \exq x \varphi(x)]}$$ 
\item[D1] A \emph{God-like} being possesses all positive properties: 
  $$\hol{G(x) \equiv \forall \varphi [P(\varphi) \imp \varphi(x)]}$$ 
\item[A3]  The property of being God-like is positive: 
  $$\hol{P(G)}$$
\item[C\phantom{1}] Possibly, a God-like being exists: $$\hol{\pos \exq x G(x)}$$
\item[A4]  Positive properties are necessarily positive: 
  $$\hol{\allq \varphi [P(\varphi) \imp \Box \; P(\varphi)]}$$ 
\item[D2] An \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties: $$\hol{\ess{\varphi}{x} \equiv \varphi(x) \wedge \allq
  \psi (\psi(x) \imp \nec \allq y (\varphi(y) \imp \psi(y)))}$$ 
\item[T2]  Being God-like is an essence of any
  God-like being: $$\hol{\allq x [G(x) \imp \ess{G}{x}]}$$
\item[D3] \emph{Necessary existence} of an individual is the necessary exemplification of all its essences: 
  $$\hol{\NE(x) \equiv \allq \varphi [\ess{\varphi}{x} \imp \nec
  \exq y \varphi(y)]}$$
\item[A5] Necessary existence is a positive property: $$\hol{P(\NE)}$$ 
\item[L1] If a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\exq x G(x) \imp \nec \exq y G(y)}$$
\item[L2] If possibly a god-like being exists, then necessarily a god-like being exists: 
  $$\hol{\pos \exq x G(x) \imp \nec \exq y G(y)} $$
%
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}

@@

\section{Bjordal's Alternative} \label{apx:Bjordal}

% \begin{figure}[t]
% \noindent \framebox[\columnwidth][r]{
% \begin{minipage}{.94\columnwidth}\small
In Bjordal's emendation $G$ (God-like) is taken as primitive and $P$ (Positive) is defined (cf. definition D).
\begin{itemize}
\item[B:D] A formulas $\phi$ is positive iff it is necessarily the case
  that anything which is God-like has the property $\phi$.
  $$\hol{P(\phi) \equiv \nec \allq x (G(x) \imp \phi(x))}$$ 
\item[B:L1] D is logically equivalent in S4 with the union of G\"odel's definition D1 and axioms A2, A3 and A4.
  $$\hol{D \biimp D1 \wedge A2 \wedge A3 \wedge A4}$$
  The proof splits into the two implication directions B:L1$^\rightarrow$ and B:L1$^\leftarrow$. B:L1$^\rightarrow$ can be further split into four single steps.
\item[B:D2] $\phi$ is a maximal composite of object $x$'s positive properties iff $x$ has $\phi$ and $\phi$ is positive and all positive properties $\psi$ which $x$ has are such that is necessarily the case that all objects which have $\phi$ also have $\psi$.
  $$\hol{MCP(\phi,x) \equiv (\phi(x) \wedge P(\phi)) \wedge \allq \psi ((\psi(x) \wedge P(\psi)) \imp \nec \allq y (\phi(y) \imp \psi(y)))}$$
\item[B:D3] $x$ has the $N$-property iff x is such that if $\phi$ is a maximal composite of $x$'s positive properties then it is necessary that some object $y$ has the property $\phi$.
  $$\hol{N(x) \equiv \allq \phi (MCP(\phi,x) \imp \nec \allq y \phi(y))}$$
\item[B:A1] If a property is positive, its negation is not positive:
  $$\hol{\allq \varphi [P(\varphi) \imp \neg P(\neg \varphi)]}$$ 
\item[B:A2] The $N$-property is positive.
 $$\hol{P(N)}$$
\item[T3] Necessarily, a God-like being exists: $$\hol{\nec \exq x G(x)}$$ 
\end{itemize}
% \end{minipage}
% } \vskip-.5em
% \caption{Frode's Alternative \cite{ToDo:Frode}.\label{fig:frode}} 
% \end{figure}
